### 2012-02-04 contents.tex
revision.

% \TODO why the VERIFIED status is so important?

@@

% \TODO Roles: reporter, fixer, verifier, reopener, closer.

% \TODO bugs with status \VERIFIED\ are used to write the release notes / change log.

@@

\begin{section}{Method} \label{sec:method}

@@

	% In order to obtain insight about the research questions, we read a sample of bug reports from both projects. For each project, two subprojects were chosen randomly among the projects with most verified bugs, resulting in the following subprojects: Eclipse/Platform, Eclipse/EMF, Netbeans/platform, Netbeans/versioncontrol. We have sampled 20 bug reports for each subproject, totalizing 80 bug reports.
	% 
	% \TODO this sampling was made just for the preliminary, qualitative analysis. Maybe we should describe preliminary analysis as an unstructured inquiry (instead of detailing the methods for it).

@@

	\begin{itemize}
		\item automated testing: testcase, test case, unit test, automated test, test result, run the test, test run;
		\item source code inspection: inspecting, inspection, look at the code, looking at the source;
		\item ad hoc testing: verified by running, verified by [verb]ing;
		\item automated static analysis: findbugs, checkstyle, static code, static analys;
		\item formal testing: formal test, format verif.
	\end{itemize}

@@

% \begin{subsection}{Preliminary Analysis}
% 	Some insights were gained in this step, partially answering the who, when, and how questions:
% 	
% 	\begin{itemize}
% 		\item [how] (OK) In some projects, the status VERIFIED does not correspond to software verification. In Eclipse/EMF, for example, it means that the fix is available in a build that is promoted on the project website\footnote{\url{http://wiki.eclipse.org/Modeling_PMC_Meeting,_2007-10-16} (between releases 3.3.1 and 3.3.1.1)}. Further evidence: bug #254489 ``Please do not change the state of the bugs since the status verified fixed should be only there when we publish a new build. :-)''
% 		\item [(who)] (NOT RELEVANT) In the NetBeans subprojects that were analyzed, it was common that the reporter also verified the fix. In some bugs, this was necessary since the developer was not able to reproduce the bug.
% 		\item [(who, when)] Often, the fixer verifies the bug just after marking it as fixed.
% 		\item [(when)] (OK) Sometimes, several old bugs are marked as VERIFIED at the same time.
% 		\item [(how?)] (OK) In both Eclipse and NetBeans, it is common that the fixer requests someone else to verify the fix (``Please verify, Tom'').
% 		\item [(how)] (OK) In Eclipse/EMF, bug #249436, one developer rewrites the patch submitted by other developer.
% 		\item [(how)] (OK) In Eclipse/EMF, bug #269789, a test case is used to assess the proposed fix.
% 		\item [(how)] (OK) Eclipse/Platform, bug #210533: ``Verified by code inspection''.
% 	\end{itemize}
% 	
% 	\TODO: ``Verified in/with [build]'', ``Verified by/through [verification technique]''.
% 	
% \end{subsection}

@@

  Most verification comments do not provide any clue about the verification technique used. Many comments just state that the bug fix was verified, optionally informing the build used to verify the fix.

	\TODO table

@@

  In Eclipse/Platform, there are 13,483 verifications, each one with a message. By the search criteria, 494 were identified as code inspections, 118 were identified as tests and 74 were identified as ad hoc.

	Evidence of both code inspection and automated testing was found in all projects, although the evidence was weaker.

	\textbf{Observations}.
  We also noted that, in Eclipse/Platform, often the fixer requests someone else to verify the fix, by asking ``please verify, [developer name]''. If the bug is reopened, then the fixer and the verifier exchange roles. That is considered good practice.

  It is difficult to identify ad hoc cases, because they are so pervasive that they are not reported, and also they are reported in many different ways.

	\TODO Eclipse/EMF: VERIFIED = in a build available in the website.

@@

	\begin{table}
		\begin{center}
		\begin{tabular}{l|r|r}
			% \toprule
			\textbf{\centering Project} & 
			\textbf{\centering QA team size} & 
			\MyHead{2.5cm}{\textbf{\% of verifications by QA team}} \\
			\hline
			Eclipse/Platform        & 4 (2.4\%)  &  1.1\% \\
			Eclipse/EMF             & 0 (0.0\%)  &  0.0\% \\
			NetBeans/Platform       & 5 (21.7\%) & 93.3\% \\
			NetBeans/VersionControl & 5 (20.8\%) & 93.2\% \\
			% \bottomrule
		\end{tabular}
		\end{center}
		\caption{Discovered QA team for all four subprojects.}
		\label{tab:qa}
	\end{table}

@@

\begin{section}{Conclusion} \label{sec:conclusion}
  The \VERIFIED\ status is used in a variety of distinct ways, not always consistent with the traditional definition of software verification ...

	Summary of the findings:
	\begin{itemize}
		\item It is difficult to detect the verification technique used in each reported verification. 
		\item Beware of cliff walls!
		\item Know your project. 
	\end{itemize}

	With the knowledge obtained from this exploratory research, we can improve and extend our previous work on the impact of independent verification on software quality. Examples of refined questions:

	\begin{itemize}
		\item Does independent verification improve software quality when it is performed by developers from the QA team?
		\item Does independent verification improve software quality when it is performed as an separate phase of the software release cycle?
		\item Does independent verification improve software quality when a particular verification technique (e.g., inspections, unit testing) is used?
	\end{itemize}

>>>>

@@

% Roles: reporter, fixer, verifier, reopener, closer.

% bugs with status \VERIFIED\ are used to write the release notes / change log.

@@

\begin{section}{Method} \label{sec:methods}

@@

	% In order to obtain insight about the research questions, we read a sample of bug reports from both projects. For each project, two subprojects were chosen randomly among the projects with most verified bugs, resulting in the following subprojects: Eclipse/Platform, Eclipse/EMF, Netbeans/platform, Netbeans/versioncontrol. We have sampled 20 bug reports for each subproject, totalizing 80 bug reports.
	%

@@

	\begin{itemize}
		\item automated testing: testcase, test case, unit test, automated test, test result, run the test, test run;
		\item source code inspection: inspecting, inspection, look at the code, looking at the source;
		\item ad hoc testing: verified by running, verified by ...ing;
		\item automated static analysis: findbugs, checkstyle, static code, static analys;
		\item formal testing: formal test, formal verif.
	\end{itemize}

@@

% \begin{subsection}{Preliminary Analysis}
% 	Some insights were gained in this step, partially answering the who, when, and how questions:
% 	
% 	\begin{itemize}
% 		\item [how] (OK) In some projects, the status VERIFIED does not correspond to software verification. In Eclipse/EMF, for example, it means that the fix is available in a build that is promoted on the project website\footnote{\url{http://wiki.eclipse.org/Modeling_PMC_Meeting,_2007-10-16} (between releases 3.3.1 and 3.3.1.1)}. Further evidence: bug #254489 ``Please do not change the state of the bugs since the status verified fixed should be only there when we publish a new build. :-)''
% 		\item [(who)] (NOT RELEVANT) In the NetBeans subprojects that were analyzed, it was common that the reporter also verified the fix. In some bugs, this was necessary since the developer was not able to reproduce the bug.
% 		\item [(who, when)] Often, the fixer verifies the bug just after marking it as fixed.
% 		\item [(when)] (OK) Sometimes, several old bugs are marked as VERIFIED at the same time.
% 		\item [(how?)] (OK) In both Eclipse and NetBeans, it is common that the fixer requests someone else to verify the fix (``Please verify, Tom'').
% 		\item [(how)] (OK) In Eclipse/EMF, bug #249436, one developer rewrites the patch submitted by other developer.
% 		\item [(how)] (OK) In Eclipse/EMF, bug #269789, a test case is used to assess the proposed fix.
% 		\item [(how)] (OK) Eclipse/Platform, bug #210533: ``Verified by code inspection''.
% 	\end{itemize}
% 	
% \end{subsection}

@@

  Most verification comments do not provide any clue about the verification technique used. Most often, developers just state that the bug fix was verified, sometimes informing the build used to verify the fix.

	Table \ref{tab:how} shows, for each subproject, the number of comments associated with a bug being marked as \VERIFIED\ or \REOPENED\ that refers to a particular verification technique. No references to formal verification or static analysis were found; therefore, these techniques are not shown on the table.

	\begin{table}
		\begin{center}
		\caption{Verification techniques for all four subprojects.}
		\begin{tabular}{l|r|r|r}
			% \toprule
			\textbf{\centering Project} & 
			\textbf{\centering Testing} & 
			\textbf{\centering Inspection} & 
			\textbf{\centering Ad Hoc} & 
			\hline
			Eclipse/Platform        & 260 (1.1\%) & 524 (2.2\%) & 74 (0.3\%) \\
			Eclipse/EMF             & 27 (0.7\%) & 2 (0.1\%) & 0 (0.0\%) \\
			NetBeans/Platform       & 94 (0.6\%) & 5 (0.0\%) & 0 (0.0\%) \\
			NetBeans/VersionControl & 4 (0.1\%) & 1 (0.0\%) & 0 (0.0\%) 
			% \bottomrule
		\end{tabular}
		\end{center}

		\label{tab:how}
	\end{table}

	In Eclipse/Platform, less than 4\% of the comments referenced the verification technique that was used; for the other projects, less than 1\% of the comments contained such references. Such low proportion reveals how difficult it is to infer the verification technique that was applied using only data from bug repositories. This is probably tacit knowledge that can only be extracted by asking the developers themselves.

	In all subprojects, comments suggest the use of automated testing and code inspection in the verification process, the former technique being more frequently referenced (except in Eclipse/Platform, in which code inspection is cited more often). Evidences of ad hoc testing were found only in Eclipse/Platform, because verification comments tend to be more structured in this project, but we believe that ad hoc testing is much more common.

	While sorting through comments, we have found interesting information regarding the software verification process. In Eclipse/Platform, the developer who fixes a bug often someone else to verify the fix, by asking ``please verify, [developer name]''. If the bug is reopened, then the fixer and the verifier exchange roles. This behavior illustrates a structured bug fixing/verification process.

	In Eclipse/EMF, we found that marking a bug as \VERIFIED\ does not mean that the bug fix was verified. Instead, it means that the fix was made available in a build of the software that is published in the subproject's website. This reinterpretation of the Bugzilla workflow was decided in a meeting\footnote{\url{http://wiki.eclipse.org/Modeling_PMC_Meeting,_2007-10-16}}.

@@

	%   In Eclipse/Platform, there are 13,483 verifications, each one with a message. By the search criteria, 494 were identified as code inspections, 118 were identified as tests and 74 were identified as ad hoc.
	% 
	% Evidence of both code inspection and automated testing was found in all projects, although the evidence was weaker.

	% \textbf{Observations}.
	%   We also noted that, in Eclipse/Platform, often the fixer requests someone else to verify the fix, by asking ``please verify, [developer name]''. If the bug is reopened, then the fixer and the verifier exchange roles. That is considered good practice.
	% 
	%   It is difficult to identify ad hoc cases, because they are so pervasive that they are not reported, and also they are reported in many different ways.

	% Eclipse/EMF: VERIFIED = in a build available in the website.

@@

	\begin{table}
		\begin{center}

		\caption{Discovered QA team for all four subprojects.}
		\begin{tabular}{l|r|r}
			% \toprule
			\textbf{\centering Project} & 
			\textbf{\centering QA team size} & 
			\MyHead{2.5cm}{\textbf{\% of verifications by QA team}} \\
			\hline
			Eclipse/Platform        & 4 (2.4\%)  &  1.1\% \\
			Eclipse/EMF             & 0 (0.0\%)  &  0.0\% \\
			NetBeans/Platform       & 5 (21.7\%) & 93.3\% \\
			NetBeans/VersionControl & 5 (20.8\%) & 93.2\% \\
			% \bottomrule
		\end{tabular}
		\end{center}

		\label{tab:qa}
	\end{table}

@@

\begin{section}{Conclusion} \label{sec:conclusion}
	By analyzing four subprojects (two from Eclipse, two from NetBeans), we have found, using only data from bug repositories, subprojects with and without QA teams, with and without a well-defined verification phase. We also have found evidence of the application of automated testing and source code inspection, although such evidence was less reliable. Also, we have found cases in which marking a bug as \VERIFIED\ does not imply that any kind of software verification was actually performed.

	With the knowledge obtained from this exploratory research, we aim to improve and extend our previous work on the impact of independent verification on software quality. We can investigate, for example, whether verification performed by QA team is better than verification performed by other developers.

	Researchers should be aware that information about verification techniques are not common in bug repositories, and that reported verification does not always correspond to actual verification. Some exploration of the data is important to avoid such pitfalls.

	% \begin{itemize}
	% 	\item Does independent verification improve software quality when it is performed by developers from the QA team?
	% 	\item Does independent verification improve software quality when it is performed as an separate phase of the software release cycle?
	% 	\item Does independent verification improve software quality when a particular verification technique (e.g., inspections, unit testing) is used?
	% \end{itemize}
